DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2022-04-07 manuscript/manuscript.Rmd
editing the text

documentclass     : "apa6"
classoption       : "man"
output            :
  papaja::apa6_pdf:
    includes      :
      in_header: "preamble.tex"
      after_body  : ["figures-alt.tex", "figures-supplementary.tex"]

@@

The DateLife algorithm is fully implemented using the R language. The latest stable version of the R package `datelife` -- v`r packageVersion("datelife")`, is available from the CRAN repository [@datelifeR], and relies on functionalities from various biological R packages:
ape [@paradis2004],
bold [@Chamberlain2018],
geiger [@Harmon2008],
paleotree [@Bapst2012a],
phyloch [@Heibl2008],
phylocomr [@Ooms2018],
phytools [@Revell2012],
rotl [@Michonneau2016], and
taxize [@Chamberlain2013; @Chamberlain2018].
Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife algorithm: providing an input, searching a chronogram database, and summarizing results from the search.

@@

_**Choosing a topology.--**_ DateLife requires a tree topology to summarize age data upon. We recommend that users provide one as input from the literature, or one of their own making. If no topology is provided, DateLife automatically subsets one from the OpenTree synthetic tree [@opentreeoflife2019synth]. Alternatively, DateLife can reconstruct a tree with branch lengths proportional to substitution rates from a starting tree topology using genetic data from the Barcode of Life Data System, BOLD [@ratnasingham2007bold], or combine topologies from source chronogras using a supertree approach.

_**Reconstructing branch lengths.--**_ DateLife starts by mining the BOLD database to obtain genetic markers for the input taxa, and aligning them with MUSCLE [@edgar2004muscle; or MAFFT, @katoh2009multiple]. Currently, branch length reconstruction is performed with parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed [@schliep2011phangorn].

@@

_**Summarizing congruified ages.--**_ Congruified pairwise node ages are used to calculate summary statistics (mean, median, minimum and maximum ages) per taxon pair. <!-- we should calculate summary statistics per node -->
<!--??? How is 'calibrations' defined here? the calibrations from congruification, or the pairwise distances??? is calibrations from congruification-->
into a single summary matrix using different methods.
Summarizing options implemented include Super Distance Matrix method [SDM, @Criscuolo2006] and summary statistics.

@@

# Case studies

@@

A college educator wishes to obtain state-of-the-art data on time of evolutionary origin of species belonging to the true finches for their class. They decide to use `datelife` because they are teaching best practices for reproducibility.
Students have the option to go to the website at www.datelife.org and perform an interactive run. However, the educator also wants the students to practice their R skills.
The first step is to run a `datelife` query using the "get species from taxon" flag. This will get all recognised species names within their chosen inclusive taxon. The Fringillidae has `r length(dquery$cleaned_names)` species, according to the Open Tree of Life taxonomy.
Once with a curated set of species taxon names, the next step is to run a `datelife` search that will find all chronograms that contain at least two species names. The algorithm proceeds to prune the trees to keep matching species names on tips only, and transform the pruned trees to pairwise distance matrices.
 There are `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
The final step is to summarize the available information using two alternative types of summary chronograms, median and SDM. As explained in the "Description" section, data from source chronograms is first summarised into a single distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig. \ref{fig:fringillidages}). Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
<!-- ????????????????? Say some things about the results!
ANSWER:
(Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
<!-- and hierarchical clustering method which appears to push ages back -->

@@

<!--Data from source chronograms can be used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates-->
To perform a cross validation analysis of the DateLife workflow, we used resulting data from the previous section (Casestudy: Real Life Application). We took individual tree topologies from each of the 19 source chronograms found (Supplementary data XX). Then we congruifued age data of source chronograms from studies, and used this ages to date the tree topology with the program BLADJ.

We found that node ages from original study, and ages estimated using all other age data available are generally correlated (Supplementary Fig. \ref{fig:cvXY}).
In 5 studies, more inclusive nodes have older original ages, and less inclusive nodes have younger original ages than their cross-validated age estimates. 
Accordingly, root ages are generally older in the original study than estimated using cross-validated ages. Root ages were similar in original and cross validated ages in three vases (Supplementary Fig. \ref{fig:cvXYdiffs}).
<!-- We need to think about the root! is it comparable across source chronograms?? not all of them have the same sampling... so probably not the same root and thus not comparable. Address this...   ANSWERED: -->
Notably, chronograms have different species sampling, hence roots are not comparable across studies. Yet, chronograms with a higher sampling number can potentially inform the age of the root of chronograms with less sampling.

<!-- 
%```{r, child = "cross-validation-pathd8.Rmd"}
%```
-->

>>>>

documentclass     : "apa6"
classoption       : "man"
output            :
  papaja::apa6_pdf:
    includes      :
      in_header: "preamble.tex"
      after_body  : "figures-alt.tex"

@@

The DateLife algorithm is fully implemented using the R language. The latest stable version of the R package `datelife` is available from the CRAN repository (v`r packageVersion("datelife")`; @datelifeR), and relies on functionalities from various biological R packages:
ape [@paradis2004],
bold [@Chamberlain2018],
geiger [@Harmon2008],
paleotree [@Bapst2012a],
phyloch [@Heibl2008],
phylocomr [@Ooms2018],
phytools [@Revell2012],
rotl [@Michonneau2016], and
taxize [@Chamberlain2013; @Chamberlain2018].
Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife algorithm: providing an input, searching a chronogram database, and summarizing results from the search.

@@

_**Choosing a topology.--**_ DateLife requires a tree topology to summarize age data upon. We recommend that users provide one as input from the literature, or one of their own making. If no topology is provided, DateLife automatically subsets one from the OpenTree synthetic tree [@opentreeoflife2019synth]. Alternatively, DateLife can reconstruct a tree with branch lengths proportional to substitution rates from a starting tree topology using genetic data from the Barcode of Life Data System [BOLD; @ratnasingham2007bold], or combine topologies from source chronogras using a supertree approach.

_**Reconstructing branch lengths.--**_ DateLife starts by mining the BOLD database to obtain genetic markers for the input taxa, and aligning them with MUSCLE [@edgar2004muscle; or MAFFT @katoh2009multiple]. Currently, branch length reconstruction is performed with parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed [@schliep2011phangorn].

@@

_**Summarizing congruified ages.--**_ Congruified pairwise node ages are used to calculate summary statistics (mean, median, minimum and maximum ages) per taxon pair. <!-- we should calculate summary statistics per node -->
<!--??? How is 'calibrations' defined here? the calibrations from congruification, or the pairwise distances??? is calibrations from congruification-->
into a single summary matrix using different methods.
Summarizing options implemented include Super Distance Matrix method [SDM, @Criscuolo2006] and .

@@

# Case study

@@

A college educator wishes to obtain state-of-the-art data on time of evolutionary origin of species belonging to the true finches for their class. They decide to use `datelife` because they are teaching best practices for reproducibility.
Students have the option to go to the website at www.datelife.org and perform an interactive run. However, the educator also wants the students to practice their R skills.
The first step is to run a `datelife` query using the "get species from taxon" flag. This will get all recognised species names within their chosen inclusive taxon. The Fringillidae has `r length(dquery$cleaned_names)` species, according to the Open Tree of Life taxonomy.
Once with a curated set of species taxon names, the next step is to run a `datelife` search that will find all chronograms that contain at least two species names. The algorithm proceeds to prune the trees to keep matching species names on tips only, and transform the pruned trees to pairwise distance matrices.
 There are `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
The final step is to summarize the available information using two alternative types of summary chronograms, median and SDM. As explained in the "Description" section, data from source chronograms is first summarised into a single distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig. \ref{fig:fringillidages}). Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
????????????????? Say some things about the results!
<!-- (Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
<!-- and hierarchical clustering method which appears to push ages back -->

@@

<!--Data from source chronograms can be used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates-->
As a form of cross validation, we took the tree topologies from each study and used time of lineage divergence data from all other source chronograms as fixed calibrations with the program BLADJ (Figs. \ref{fig:cv1} to \ref{fig:cv10}).
<!--I'm confused about why you are talking about branch lengths here... did you use branch lengths? Yes, on the PATHd8 cross validation, but you are right, it is confusing, so I removed that first sentence-->

The ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvbladj}).
Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cv2}).
<!-- We need to think about the root! is it comparable across source chronograms?? not all of them have the same sampling... so probably not the same root and thus not comparable. Address this -->
We also demonstrate the usage of PATHd8 [@Britton2007] as a dating method alternative to BLADJ. PATHd8 requires a tree with initial branch lengths which we obtained using `datelife`'s branch length reconstruction algorithm, that searches for DNA sequence data from the Barcode of Life Data System [BOLD; @ratnasingham2007bold] to generate branch lengths. We were able to successfully generate a tree with BOLD branch lengths for all of the Fringillidae source chronograms. However, dating with PATHd8 using congruified calibrations, was only successful in
<!-- two studies when expanding calibrations to make them agree (studies 3 and 5), and from  -->
three cases (studies 3, 5, and 9, shown in Fig. \ref{fig:cvbold}).
<!-- What made it unsuccessful in the other cases? -->
From these, two trees have a different sampling than the original source chronogram, mainly because DNA BOLD  data for some species is absent from the database.
<!-- ??? Node ages or maximum ages?? -->
Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
More examples and code used to generate these trees were developed on an open repository that is available for consultation and reuse at <https://github.com/LunaSare/datelife_examples>.

### 2022-04-07 manuscript/manuscript.tex
editing the text

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{DateLife: revealing the dated Tree of Life}
\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4201}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

A DateLife search consists of matching processed taxon names to tip labels in a chronogram database. Chronograms with at least two matching tip labels are identified and pruned down to preserve only the matched tips.

Matching pruned chronograms are stored as individual patristic distance matrices (Figure \ref{fig:workflow} subfigure X).
This matrix consists of \ldots{} ???? the pairwise distance between pairs of query taxa which are in that input tree, in units of millions of years.

This format speeds up extraction of pairwise taxon ages of the queried taxa, as opposed to searching the ancestor node of a pair of taxa in a ``phylo'' object or newick string.
The patristic matrices are also associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.

DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies. It is constructed from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert phylogenetic knowledge with rich metadata (McTavish et al., 2015) that allows automatic and reproducible construction of a chronogram database.
New chronograms can be added to Phylesystem by any user and are immediately publicly available, and the DateLife database can be updated to include those new data within a run.

@@

At this point, summary information is extracted from the \texttt{datelifeResult} object to inform decisions for the subsequent steps in the user workflow. Age data from the matching pruned chronograms is summarized and used to generate a single summary chronogram. Other basic summary information available to the user is:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The matching pruned chronograms as newick strings or ``phylo'' objects.
\item
  The ages of the root of all matching pruned chronograms. This can correspond to the age of the most recent common ancestor (mrca) of your group of interest if the pruned chronograms have all taxa belonging to the group. If not, the root corresponds to the mrca of a subgroup withing your group of interest.
\item
  Study citations where original chronograms were published.
\item
  A report of input taxon names matches across pruned chronograms.
\item
  The single matching pruned chronogram with the most input taxon names.
\end{enumerate}

\emph{\textbf{Identifying groves.--}} To generate a single summary chronogram, the DateLife algorithm starts by identifying the matching pruned chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al. (2009). In rare cases, a group of trees can have multiple groves. By default, DateLife chooses the grove with the most taxa, however, the ``criterion = trees'' flag allows the user to choose the grove with the most trees instead.

\emph{\textbf{Choosing a topology.--}} DateLife requires a tree topology to summarize age data upon. Users can provide one as input from the literature, or one of their own making. If no topology is provided, DateLife automatically subsets one from the OpenTree synthetic tree (Open Tree Of Life et al., 2019).

DateLife can also reconstruct branch lengths proportional to substitution rates on a fixed tree topology using available genetic data from BOLD.

\emph{\textbf{Congruifying nodes.--}} DateLife then implements the congruification method (Eastman, Harmon, \& Tank, 2013) to find nodes belonging to the same clade across matching pruned chronograms.
Congruified node ages stored as a \texttt{congruifiedCalibrations} object are then matched to nodes in the chosen tree topology and stored as a \texttt{matchedCalibrations} object.

\emph{\textbf{Summarizing node ages.--}} DateLife summarizes matched calibrations
into a single patristic distance matrix using different methods.
Summarizing options implemented include Super Distance Matrix method (SDM, Criscuolo, Berry, Douzery, \& Gascuel, 2006) and summary statistics such as median, minimum and maximum ages.

\emph{\textbf{Dating the tree topology.--}} Summarized calibrations can be applied as secondary calibrations with different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Campbell O. Webb, Ackerly, \& Kembel, 2008; Campbell O. Webb \& Donoghue, 2005), and treePL (Stephen A. Smith \& O'Meara, 2012).

By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm that assigns ages to nodes with no data evenly between nodes with age data, which minimizes age variance in the resulting chronogram (Campbell O. Webb et al., 2008).
When there is conflict in ages across node with age data, the algorithm ignores ages that are older than parent nodes and/or younger than descendant nodes.

@@

@@

\texttt{datelife}'s code speed was tested on an Apple iMac
with one 3.4 GHz Intel Core i5 processor.
We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
Query processing time increases roughly linearly with number of input taxon names, and
increases considerably if TNRS is activated.
Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig. \ref{fig:runtime_main}).
\texttt{datelife}'s code performance was evaluated with a set of unit tests designed and
implemented with the R package testthat (R Core Team, 2018) that were run both locally
with the devtools package (R Core Team, 2018), and on a public server --via
GitHub, using the continuous integration tool Travis CI (\url{https://travis-ci.org}). At
present, unit tests cover more than 40\% of \texttt{datelife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).

@@

We chose 6 bird species associated to true finches at random. The sample includes
two species of cardinals:
the black-thighed grosbeak -- \emph{Pheucticus tibialis} and
the crimson-collared grosbeak -- \emph{Rhodothraupis celaeno};
three species of buntings:
the yellowhammer -- \emph{Emberiza citrinella},
the pine bunting -- \emph{Emberiza leucocephalos} and
the yellow-throated bunting -- \emph{Emberiza elegans};
and one species of tanager, the vegetarian finch -- \emph{Platyspiza crassirostris}.

Processing input names found that \emph{Emberiza elegans} is synonym for \emph{Schoeniclus elegans} in the reference taxonomy.
DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies. Three studies matched five input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2015; Hedges, Marin, Suleski, Paymer, \& Kumar, 2015; Jetz, Thomas, Joy, Hartmann, \& Mooers, 2012), one study matched four input names (Hooper \& Price, 2017) and two studies matched two input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2013; Burns et al., 2014). No studies matched all input names. Together, matching chronograms have 28 unique age data points. All nodes have age data.
As fixed tree topology, DateLife used OpenTree's synthetic tree as default and mapped age data to nodes in the tree.
As expected, more inclusive nodes (e.g., node ``n1'') have more age data than less inclusive nodes (e.g., node ``n5'').
The processing step allowd discovering five data points for node ``n4'' that would not have had any data otherwise.
Age summary statistics per node were calculated and tested as secondary calibrations to date the tree topology using the BLADJ algorithm.
Age data for node ``n2'' was excluded as final calibration because it is older than age data of a more inclusive node.

@@

Data from source chronograms can be used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates (Figs. \ref{fig:cv1} to \ref{fig:cv10}). As a form of cross validation, we took tree topologies from each input study and calibrated them using time of lineage divergence data from all other source chronograms.

In the absence of branch lengths, the ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvbladj}).
Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cv2}).
We also demonstrate the usage of PATHd8 (Britton et al., 2007) as an dating method alternative to BLADJ. For this, we run a \texttt{datelife} branch length reconstruction that searches for DNA sequence data from the Barcode of Life Data System {[}BOLD; Ratnasingham and Hebert (2007){]} to generate branch lengths. We were able to successfully generate a tree with BOLD branch lengths for all of the Fringillidae source chronograms. However, dating with PATHd8 using congruified calibrations, was only successful in
three cases (studies 3, 5, and 9, shown in Fig. \ref{fig:cvbold}).
From these, two trees have a different sampling than the original source chronogram, mainly because DNA BOLD data for some species is absent from the database.
Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
More examples and code used to generate these trees were developed on an open repository that is available for consultation and reuse at \url{https://github.com/LunaSare/datelife_examples}.

@@

At the time of writing of this manuscript (Apr 04, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015). A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time. As chronograms are added to Phylesystem, they are incorporated into an updated \texttt{datelife}'s database that is assigned a new version number, followed by a package release on CRAN. \texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
Users can also upload new chronograms to OpenTree themselves, and trigger an update of their local \texttt{datelife} database to incorporate the new chronograms, to have them immediately available for analysis.

@@

@@

@@

@@

@@

>>>>

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{DateLife: revealing the dated Tree of Life}
\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4379}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

A DateLife search consists of matching processed taxon names to tip labels in a chronogram database. Chronograms with at least two matching tip labels are identified and pruned down to preserve only the matched tips.
These matching pruned chronograms are now referred to as source chronograms.
Total distance (in units of millions of years) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Figure \ref{fig:workflow}).
The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a ``phylo'' object or newick string.
Finally, the patristic matrices are associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.

DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies. It is constructed from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert phylogenetic knowledge with rich metadata (McTavish et al., 2015) that allows automatic and reproducible construction of a chronogram database.
New chronograms can be added to Phylesystem by any user and are immediately publicly available. This permits an immediate update of DateLife's chronogram database to include those new data within a run.

@@

Summary information is extracted from the \texttt{datelifeResult} object to inform decisions for the subsequent steps in the user workflow. Basic summary information available to the user is:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The matching pruned chronograms as newick strings or ``phylo'' objects.
\item
  The ages of the root of all source chronograms. This can correspond to the age of the most recent common ancestor (mrca) of your group of interest if the source chronograms have all taxa belonging to the group. If not, the root corresponds to the mrca of a subgroup withing your group of interest.
\item
  Study citations where original chronograms were published.
\item
  A report of input taxon names matches across source chronograms.
\item
  The source chronogram(s) with the most input taxon names.
\item
  Last but not least, age data from source chronograms are summarized and used to generate a single summary chronogram:
\end{enumerate}

\emph{\textbf{Choosing a topology.--}} DateLife requires a tree topology to summarize age data upon. We recommend that users provide one as input from the literature, or one of their own making. If no topology is provided, DateLife automatically subsets one from the OpenTree synthetic tree (Open Tree Of Life et al., 2019). Alternatively, DateLife can reconstruct a tree with branch lengths proportional to substitution rates from a starting tree topology using genetic data from the Barcode of Life Data System {[}BOLD; Ratnasingham and Hebert (2007){]}, or combine topologies from source chronogras using a supertree approach.

\emph{\textbf{Reconstructing branch lengths.--}} DateLife starts by mining the BOLD database to obtain genetic markers for the input taxa, and aligning them with MUSCLE (Edgar, 2004; or MAFFT Katoh, Asimenos, \& Toh, 2009). Currently, branch length reconstruction is performed with parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed (Schliep, 2011).

\emph{\textbf{Combining source chronograms.--}} To combine topologies from source chronograms into a single summary topology, the DateLife algorithm starts by identifying the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al. (2009). In rare cases, a group of trees can have multiple groves. By default, DateLife chooses the grove with the most taxa, however, the ``criterion = trees'' flag allows the user to choose the grove with the most trees instead.
If source chronograms do not form a grove, the supertree reonstruction will fail.

\emph{\textbf{Congruifying nodes.--}} Once with a chosen topology, DateLife applies the congruification method (Eastman, Harmon, \& Tank, 2013) to find nodes belonging to the same clade across source chronograms, and extract the corresponding node ages from the patristic distance matrices stored as \texttt{datelifeResult}. Since by definition the matrices store total distance (time from tip to tip), node ages correspond to half the values stored in the patristic distance matrices.
A table of congruified node ages that can be used as secondary calibrations is stored as a \texttt{congruifiedCalibrations} object.

\emph{\textbf{Summarizing congruified ages.--}} Congruified pairwise node ages are used to calculate summary statistics (mean, median, minimum and maximum ages) per taxon pair.
into a single summary matrix using different methods.
Summarizing options implemented include Super Distance Matrix method (SDM, Criscuolo, Berry, Douzery, \& Gascuel, 2006) and .

Ages per taxon pair can be used as calibrations to date the chosen tree topology.

\emph{\textbf{Dating the tree topology.--}}
By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully datd topology. BLADJ fixes node ages that have calibration data, and distrivutes time between nodes with no data evenly between nodes with calibration data. This minimizes age variance in the resulting chronogram (Campbell O. Webb, Ackerly, \& Kembel, 2008).
When there is conflict in ages between nodes with calibration data, the algorithm ignores ages that are older than ages of parent nodes and/or younger than ages from descendant nodes.

@@

Summarized calibrations can be applied as secondary calibrations with different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Campbell O. Webb et al., 2008; Campbell O. Webb \& Donoghue, 2005), and treePL (Stephen A. Smith \& O'Meara, 2012).

@@

\texttt{datelife}'s code speed was tested on an Apple iMac
with one 3.4 GHz Intel Core i5 processor.
We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
Query processing time increases roughly linearly with number of input taxon names, and
increases considerably if Taxonomic Name Resolution Service (TNRS) is activated.
Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig. \ref{fig:runtime_main}).
\texttt{datelife}'s code performance was evaluated with a set of unit tests designed and
implemented with the R package testthat (R Core Team, 2018) that were run both locally
with the devtools package (R Core Team, 2018), and on a public server --via
GitHub, using the continuous integration tool Travis CI (\url{https://travis-ci.org}). At
present, unit tests cover more than 40\% of \texttt{datelife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).

@@

We randomly chose 6 bird species related to the family Fringillidae of true finches. The sample includes
two species of cardinals:
the black-thighed grosbeak -- \emph{Pheucticus tibialis} and
the crimson-collared grosbeak -- \emph{Rhodothraupis celaeno};
three species of buntings:
the yellowhammer -- \emph{Emberiza citrinella},
the pine bunting -- \emph{Emberiza leucocephalos} and
the yellow-throated bunting -- \emph{Emberiza elegans};
and one species of tanager, the vegetarian finch -- \emph{Platyspiza crassirostris}.

Processing input names found that \emph{Emberiza elegans} is synonym for \emph{Schoeniclus elegans} in the default reference taxonomy {[}Open Tree of Life Taxonomy v3.3, June 1, 2021{]}. For a detailed discussion on the state of the synonym refer to Avibase (Avibase, 2022; Lepage, 2004; Lepage, Vaidya, \& Guralnick, 2014).
DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies. Three studies matched five input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2015; Hedges, Marin, Suleski, Paymer, \& Kumar, 2015; Jetz, Thomas, Joy, Hartmann, \& Mooers, 2012), one study matched four input names (Hooper \& Price, 2017) and two studies matched two input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2013; Burns et al., 2014). No studies matched all input names. Together, matching chronograms have 28 unique age data points. All nodes have age data.
As fixed tree topology, DateLife used OpenTree's synthetic tree as default and mapped age data to nodes in the tree.
As expected, more inclusive nodes (e.g., node ``n1'') have more age data than less inclusive nodes (e.g., node ``n5'').
The processing step allowed discovering five data points for node ``n4'' that would not have had any data otherwise.
Age summary statistics per node were calculated and tested as secondary calibrations to date the tree topology using the BLADJ algorithm.
Age data for node ``n2'' was excluded as final calibration because it is older than age data of a more inclusive node.

@@

As a form of cross validation, we took the tree topologies from each study and used time of lineage divergence data from all other source chronograms as fixed calibrations with the program BLADJ (Figs. \ref{fig:cv1} to \ref{fig:cv10}).

The ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvbladj}).
Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cv2}).
We also demonstrate the usage of PATHd8 (Britton et al., 2007) as a dating method alternative to BLADJ. PATHd8 requires a tree with initial branch lengths which we obtained using \texttt{datelife}'s branch length reconstruction algorithm, that searches for DNA sequence data from the Barcode of Life Data System {[}BOLD; Ratnasingham and Hebert (2007){]} to generate branch lengths. We were able to successfully generate a tree with BOLD branch lengths for all of the Fringillidae source chronograms. However, dating with PATHd8 using congruified calibrations, was only successful in
three cases (studies 3, 5, and 9, shown in Fig. \ref{fig:cvbold}).
From these, two trees have a different sampling than the original source chronogram, mainly because DNA BOLD data for some species is absent from the database.
Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
More examples and code used to generate these trees were developed on an open repository that is available for consultation and reuse at \url{https://github.com/LunaSare/datelife_examples}.

@@

At the time of writing of this manuscript (Apr 07, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015). A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time. As chronograms are added to Phylesystem, they are incorporated into an updated \texttt{datelife}'s database that is assigned a new version number, followed by a package release on CRAN. \texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
Users can also upload new chronograms to OpenTree themselves, and trigger an update of their local \texttt{datelife} database to incorporate the new chronograms, to have them immediately available for analysis.

@@

\leavevmode\vadjust pre{\hypertarget{ref-avibase-emberiza}{}}%
Avibase. (2022). {Yellow-throated Bunting}. \emph{{Avibase - The World Bird Database}}, (Online Resource). Retrieved from \url{\%7Bhttps://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\%7D}

@@

\leavevmode\vadjust pre{\hypertarget{ref-edgar2004muscle}{}}%
Edgar, R. C. (2004). MUSCLE: Multiple sequence alignment with high accuracy and high throughput. \emph{Nucleic Acids Research}, \emph{32}(5), 1792--1797.

@@

\leavevmode\vadjust pre{\hypertarget{ref-katoh2009multiple}{}}%
Katoh, K., Asimenos, G., \& Toh, H. (2009). Multiple alignment of DNA sequences with MAFFT. In \emph{Bioinformatics for DNA sequence analysis} (pp. 39--64). Springer.

@@

\leavevmode\vadjust pre{\hypertarget{ref-lepage2004avibase}{}}%
Lepage, D. (2004). \emph{Avibase: The world bird database}. Bird Studies Canada.

\leavevmode\vadjust pre{\hypertarget{ref-lepage2014avibase}{}}%
Lepage, D., Vaidya, G., \& Guralnick, R. (2014). Avibase--a database system for managing and organizing taxonomic concepts. \emph{ZooKeys}, (420), 117.

@@

\leavevmode\vadjust pre{\hypertarget{ref-schliep2011phangorn}{}}%
Schliep, K. P. (2011). Phangorn: Phylogenetic analysis in r. \emph{Bioinformatics}, \emph{27}(4), 592--593.
